Interactive Data Exploration with Smart Drill-Down (Extended Version)

### 2016-12-02 TableSummarization/journalVersion/smartDrillDown.tex
Added content based on reviewer comments

\subsection{NP-Hardness for Problem~\ref{prob:optimal-rule-set}}
We reduce the well known {\sc NP-Hard} {\em Maximum Coverage Problem} (MCP) to a special case of Problem~\ref{prob:optimal-rule-set};
thus demonstrating the {\sc NP-Hard}ness of Problem~\ref{prob:optimal-rule-set}. MCP is given below: 
\begin{problem}\label{prob:maximum-coverage}
Given a universe set $U$, an integer $k$, and a set $S = \left\lbrace S_1, S_2, ... S_m \right\rbrace$ of subsets of $U$ (so each $S_i \subset U$), find $S^{\prime} \subset S$ such that $|S^{\prime}| = k$, which maximizes $\text{Coverage}(S^{\prime}) = |\bigcup_{s \in S^{\prime}} s|$.
\end{problem}
Thus, the goal of MCP is to find a set of $k$ of the given subsets of $U$ whose union `covers' as much of $U$ as possible. We can reduce an instance of MCP (with parameters $U, k, S$) to an instance of Problem~\ref{prob:optimal-rule-set}, which gives us the following lemma:
\begin{lemma}
Problem~\ref{prob:optimal-rule-set} is {\sc NP-Hard}.
\end{lemma}

@@

\begin{lemma}\label{lemma:submodular}
For a given table $T$, the Score function over sets $S$ of rules, defined by the following is submodular:
$$\text{Score}(S) = \sum_{r \in S} MCount(r,S)W(r)$$
\end{lemma}
\vspace{-10pt}

@@

@@

@@

Then next step is to combine the solutions using dynamic programming. Let the $M$ sets be called $M_0, M_1, ... M_D$. Let our possible sample sizes range from $0$ to $\mathcal{S}$. The number of sample sizes can be pretty large ($\mathcal{S}$), but we can make it smaller by discretizing the sample sizes, say to have granularity $100$. Then we create a $D \times \mathcal{S}$ array $A$. The value $A\left[i\right]\left[j\right]$ contains the best probability score we can get from $M_0, M_1, ... M_i$ with total sample size at most $j$. We can populate $A\left[0\right]\left[j\right] \forall j$ using the locally optimal solutions for $M_0$. Let $E_{i+1}$ denote the set of locally optimal solutions for $M_{i+1}$. Then we have,
$$A\left[i+1\right] \left[j \right] = \textrm{max} (A\left[i\right]\left[j\right], \textrm{max}_{e \in E_{i+1}}(A\left[i\right]\left[j-S(e)\right] + P(e)))$$
Dynamic programming solves this in $O(D\mathcal{S}3^d)$ time. 
}

@@

\begin{figure}
\centering
\includegraphics[width=80mm,frame]{graphs/screenshots6col/bits_scoring.jpg}
\caption{Bits scoring\label{fig:weigtingbitwise}}
\vspace{-15pt}
\end{figure}

>>>>

\subsection{NP-Hardness for Problem~\ref{prob:optimal-rule-set}}
We reduce the well known {\sc NP-Hard} {\em Maximum Coverage Problem} (MCP) to a special case of Problem~\ref{prob:optimal-rule-set};
thus demonstrating the {\sc NP-Hard}ness of Problem~\ref{prob:optimal-rule-set}. MCP is given below: 
\begin{problem}\label{prob:maximum-coverage}
Given a universe set $U$, an integer $k$, and a set $S = \left\lbrace S_1, S_2, ... S_m \right\rbrace$ of subsets of $U$ (so each $S_i \subset U$), find $S^{\prime} \subset S$ such that $|S^{\prime}| = k$, which maximizes $\text{Coverage}(S^{\prime}) = |\bigcup_{s \in S^{\prime}} s|$.
\end{problem}
Thus, the goal of MCP is to find a set of $k$ of the given subsets of $U$ whose union `covers' as much of $U$ as possible. We can reduce an instance of MCP (with parameters $U, k, S$) to an instance of Problem~\ref{prob:optimal-rule-set}, which gives us the following lemma:
\begin{lemma}
Problem~\ref{prob:optimal-rule-set} is {\sc NP-Hard}.
\end{lemma}
%%%%%%%%%%%%%%%%%%%%
\begin{proof}
Consider a table with $|U|$ rows (one per element of $U$) and $m$ columns (one per $S_j$, that has value $1$ in row $i$, column $j$ if the $i^{th}$ belongs to $S_j$ and $0$ otherwise. And consider a weighting function with $W(r) = 1$ if there is at least one $1$ in $r$, and $0$ otherwise. Let $k \leq m$.

Then a rule with multiple $1$s is clearly dominated (or matched) by a subrule with only one $1$. Moreover, a rule-list that has two rules with a $1$ in the same column is dominated by one that that $1$s in different columns. Then the score of a rule list that has rules with $1$s in column set $C \subset \{1, 2,\ldots m\}$ with $|C| = k$ has score equal to the size of union $\bigcup_{c \in C} S_c$. Thus maximizing score is equivalent to maximizing the size of the union of $k$ sets, which is the Maximum Coverage Problem. Thus the NP-Hardness of Maximum Coverage implies that Score maximization is NP-hard.
\end{proof}
%%%%%%%%%%%%%%%%%%%%

@@

\begin{lemma}\label{lemma:submodular}
For a given table $T$, the Score function over sets $S$ of rules, defined by the following is submodular:
$$\text{Score}(S) = \sum_{r \in S} MCount(r,S)W(r)$$
\end{lemma}
\papertext{\vspace{-10pt}}
\techreporttext{
\begin{proof}
Let $R(S)$ be the rule list obtained by sorting $S$ in descending weight order. For each tuple $t$, let $TOP(t,S)$ denote the first rule in rule-list $R(S)$ that covers $t$. By definition of $R(S)$, $r$ must also have the highest weight out of all rules in $S$ that cover $t$. Then $MCount(r,S)$ is simply $\sum_{t \in T : TOP(t,S) = r} 1$. Thus Score can be rewritten as:
$$\text{Score}(S) = \sum_{t \in T} W(TOP(t,S))$$
Now let $S \subsetneq S'$ and $s \notin S$. Then to prove submodularity, we simply need to prove that $$\text{Score}(S \cup \{s\}) - \text{Score}(S) \geq \text{Score}(S' \cup \{s\}) - \text{Score}(S')$$. We have
$\text{Score}(S \cup \{s\}) - \text{Score}(S) = \sum_{t \in T} W(TOP(t,S \cup \{s\})) - W(TOP(t,S))$ 
and an analogous equation for $S'$. Consider two cases: 
\begin{enumerate}
\item $W(TOP(t,S' \cup \{s\})) - W(TOP(t,S')) > 0$: This means $s$ covers $t$ and has higher weight than any rule in $S'$ that covers $t$. That is, $TOP(t,S' \cup \{s\}) = s$. Since $S \subseteq S'$, we must have $W(TOP(t,S')) \geq W(TOP(t,S))$, and also $TOP(t,S \cup \{s\}) = s$. Thus we have $W(TOP(t,S' \cup \{s\})) - W(TOP(t,S')) \leq W(TOP(t,S \cup \{s\})) - W(TOP(t,S))$.
\item $W(TOP(t,S' \cup \{s\})) - W(TOP(t,S')) = 0$: In this case, clearly 
$W(TOP(t,S' \cup \{s\})) - W(TOP(t,S')) \leq W(TOP(t,S \cup \{s\})) - W(TOP(t,S))$ since the latter is non-negative. 
\end{enumerate}
Either way, we have $W(TOP(t,S' \cup \{s\})) - W(TOP(t,S')) \leq W(TOP(t,S \cup \{s\})) - W(TOP(t,S))$, and summing over all $t \in T$ gives us 
$$\text{Score}(S \cup \{s\}) - \text{Score}(S) \geq \text{Score}(S' \cup \{s\}) - \text{Score}(S')$$
as required.
\end{proof}
}

@@

%%%%%%%%%%%%%%%%%%%%%%%
\textbf{Runtime analysis:}
Our algorithm runs calls the function for finding best marginal rule $k$ times. Thus the total computational cost is about $k$ times the cost of computing the best marginal rule. Let $c_n$ be the number of counts in $C_n$ at the end of the $n^{th}$ iteration (that is, after eliminating rules that have no chance of having the highest marginal value). Also, let $c_0 = 1$.

The cost of the finding the best marginal rule can be split into three parts: (i) Enumerating all new rules to add to $C_n$ at the start of the $n^{th}$ iteration. (ii) Determining which of the rules in $C_n$ to eliminate (iii) Finding counts and marginal values for the remaining $c_n$ rules in $C_n$. The number of super-rules we can consider for (i) can be no larger than $c_{n-1}|T|$, since each rule in $C_{n-1}$ can give us at most $|T|$ new candidate super-rules for $C_n$. The cost of (ii) is $O(|C_n|)$. Finally, the cost of (iii) is again $c_n|T|$, since we have two nested for loops over $T$ and $C_n$ respectively. Thus our total cost is $O(|T|\sum_i c_i)$.

In the worst case, we may need to evaluate all possible rules; since the total possible number of rules is $|T||C|$ where $C$ is the set of columns, our cost is at most $O(k|C||T|^2))$. However, note that this worst case is unlikely to occur when we use the Size or Bits weighting function, because the count of a rule decreases rapidly with its size, making it unlikely that we will need to find counts for its super-rules.

In fact, if we assume that values in columns occur independently of each other, and that the most frequent value occurs $x < 1$ fraction of the time, then the number of candidate rules is bounded by $c_i \leq c_1 x^i$. This gives us $\sum_i c_i \leq c_1\sum_i x^i \leq c_1/(1-x)$.
Thus our cost is now only $O(k|T|c_1/(1-x))$ for some constant $x$. If the number of distinct values in the table ($c_1$), $k$ and $x$ are taken to be constants (i.e. not scale with table size), then our cost is $O(|T|)$.

\textbf{Approximation ratio:}
If our $m_w$ is higher than the actual weight of every rule in the best rule list, then the only approximation ratio we incur is that in the greedy approximation, which is a multiplicative factor of up to $1 - 1/e$. On the other hand, if our $m_w$ estimate is smaller than the actual maximum weight $m_w^*$ of a rule in the best list, then we may end up picking a sub-optimal rule in the best marginal rule function.

We now bound the sub-optimality of our chosen rule. Suppose the best rule to pick is $r^*$, and we pick rule $r$. Since we estimated $r^*$'s weight to be at most $m_w$, we would have estimated $r^*$ to have marginal value $\text{Est}(r^*) = \sum_{t \in r^*}(m_w - w(t))$ where $t \in r^*$ iterates over all tuples covered by $r^*$, and $w(t)$ is the weight of the current highest weight rule covering $t$. Since we picked $r$ over $r^*$, it means $\text{MarginalValue}(r) \geq \text{Est}(r^*)$. Suppose $m$ is the highest weight of a rule selected in the previous steps of BRS (so $w(t) \leq m$ for all $t$). Then the actual marginal value of $r^*$ is $\leq \frac{m_w - m}{m_w^* - m}\text{Est}(r^*)$. Thus our pick of $r$ is at most a $\frac{m_w - m}{m_w^* - m}$ factor worse than the optimal pick. If we always choose  our $m_w$ such we have $\frac{m_w - m}{m_w^* - m} \geq w$, for some constant $w < 1$, then our final score is guaranteed to be withing a $w(1-1/e)$ factor of the optimal score.
%%%%%%%%%%%%%%%%%%%%%%%

@@

%%%%%%%%%%%%%%%%%%
While the setting of $W$ (and hence $m_w$ is mostly specific to the user, we will consider a simple model of how frequently values occur in a table and determine a weighting function to favor rules of a particular size based on that model.

Let us consider a parametric weight function which generalizes the `Bits Weighting Function' from Section~\ref{sec:weighting}. 
$W(r) = \left(\sum_{c \in C} o_{r,c}w_c\right)^k$, where the $w_c$'s and $k$ are parameters of the weighting function, and $o_{r,c}$ ($o$ stands for occurrence) is $1$ if $r(c) \neq \star$ and $0$ otherwise. The Size and Bits weighting functions can be seen as special cases of the above function with $k=1$ and with $w_c$ equal to $1$ for Size and $\text{log}(|c|)$ for Bits.

We want to estimate the Score contribution of a rule according to this weight function, which requires estimating the count of the rule. Suppose for each column $c$, the most frequent value occurs $f_c$ fraction of the time. If the probability of a value occurring in a column of a tuple is independent of other values in the tuple, the $Count$ of rule $r$ is approximately $|T|\Pi_{c \in C}f_c^{o_{r,c}}$. Thus the score of a rule set containing $r$ alone can be approximated by:
$$S(r) = \left(\sum_{c \in C} o_{r,c}w_c\right)^k|T|\Pi_{c \in C}f_c^{o_{r,c}}$$

Now consider the rule that has the highest weight times count according to the above weighting function. We can approximate this rule by relaxing $o_{r,c}$ to a real number; that is by maximizing the estimated $S(r)$ subject to $0 \leq o_{r,c} \leq 1$. According to the KKT conditions, at the optimum, the partial derivative of $W(r)$ with respect to $o_{r,c}$ must be $\leq 0$ for each $c$ where $o_{r,c} = 0$, equal to $0$ if $0 < o_{r,c} < 1$ and $\geq 0$ otherwise. For each $c$, we have:
$$\frac{\partial S(r)}{\partial o_{r,c}} = S(r)(\frac{kw_c}{\sum_{c \in C} o_{r,c}w_c} + \ln f_c)$$

Thus if $x = \frac{k}{\sum_{c \in C} o_{r,c}w_c}$, then the maximum scoring rule should instantiate columns where $\frac{\ln f_c}{w_c} > x$ and some of the ones with $\frac{\ln f_c}{w_c} = x$. Thus for Size weighting, the highest score rule should select columns with high values of $\ln f_c$ i.e. highest frequency values. In contrast, Bits will select columns with highest values of $\log_{|c|}(f_c)$ i.e. values that are frequent relative to the number of distinct values in the column.

If we want the score to treat all columns equally, then the derivative must be $0$ for all $c$. This implies that for all $c$, we must have $w_c \propto \ln f_c$. Notice that Bits satisfies the above criterion if we assume that values in each column are uniformly distributed (that is, $f_c = 1/|c|$). This provides an intuitive justification for we might want to use the Bits weighting function.

We can also estimate the size of the optimal rule under the above weighting function. Specifically, the weighted fraction of columns with non-$\star$ values is given by $\frac{\sum_{c\in C} o_{r,c}w_c}{\sum_{c \in C} w_c}$ which equals $-k/(\sum_{c \in C} \ln f_c)$. Thus if we wanted our highest score rule to have $s$ fraction of the columns get instantiated, we can achieve that by setting $k = -s\sum_{c \in C} \ln f_c$.

The above reasoning also lets us estimate the $m_w$ parameter we need to use. Specifically, the highest score rule above has weight $(\frac{-\ln f_c}{kw_c})^k$.

We can use the above estimate to guess the size of the max scoring rule for the Bits weighting function. But note that in practice, values are not distributed uniformly (so $f_c$ is actually higher than $1/|c|$) and are not independent of each other, due to correlations. As a result, our estimate of the maximum size is a significant under-estimate. In our experiments (Section~\ref{sec:exp_qualitative}), we observe that use the Bits weighting function actually gives us rules of size $3$ on our datasets, while our estimate above is approximately $0.74$. As expected, the values displayed by the rules are ones that occur much more frequently relative to other values in that column.

%This evaluates to $1.44$, while the observed value in our experiments is $10$. 
%%%%%%%%%%%%%%%%%%

@@

The% next step is to combine the solutions using dynamic programming. Let the $M$ sets be called $M_0, M_1, ... M_D$. Let our possible sample sizes range from $0$ to $\mathcal{S}$. The number of sample sizes can be pretty large ($\mathcal{S}$), but we can make it smaller by discretizing the sample sizes, say to have granularity $100$. Then we create a $D \times \mathcal{S}$ array $A$. The value $A\left[i\right]\left[j\right]$ contains the best probability score we can get from $M_0, M_1, ... M_i$ with total sample size at most $j$. We can populate $A\left[0\right]\left[j\right] \forall j$ using the locally optimal solutions for $M_0$. Let $E_{i+1}$ denote the set of locally optimal solutions for $M_{i+1}$. Then we have,
$$A\left[i+1\right] \left[j \right] = \textrm{max} (A\left[i\right]\left[j\right], \textrm{max}_{e \in E_{i+1}}(A\left[i\right]\left[j-S(e)\right] + P(e)))$$
Dynamic programming solves this in $O(D\mathcal{S}3^d)$ time. 
}

@@

\begin{figure}
\centering
\includegraphics[width=80mm,frame]{graphs/screenshots6col/bits_scoring.jpg}
\caption{Bits scoring\label{fig:weigtingbitwise}}
\end{figure}
